Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-11 Response.tex

\setcounter{page}{1}

\bigskip
\medskip
\noindent{\Large 
Response 
}
\bigskip

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%---SUGGESTED SIMULATION SCENARIOS---%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section*{Simulations}
% 
% \begin{itemize}
% 
% \subsection*{David Posada}
% 
% \item {
% {\it
% \textbf{
% Simulate a number of epochs that do not correspond with the assumed ones.
% }% END: textbf
% }% END: it
% }% END: item
% \\
% Running. \\
% Simulate 3, infer 2 : simulationSet2 \\
% Simulate 2 infer 3 : simulationSet1 
% 
% 
% \item {
% {\it
% \textbf{
% Simulating under a non-epoch model, for example where different process can apply to contemporaneous branches, would be fundamental to understand the robustness of the model.
% }% END: textbf
% }% END: it
% }% END: item
% \\
% Running (trend\_simulation).
% 
% 
% 
% \item {
% {\it
% \textbf{
% Here it would have been interesting to see the correspondence between simulated and estimated tMRCAs when the true model has one epoch with omega=0.5 (or even better, omega equal to the weighted mean of the n-epoch model).
% }% END: textbf
% }% END: it
% }% END: item
% \\
% Running (deepRootv2).

% \subsection*{Stijn Vansteelandt}
% 
% \begin{itemize}
% 
% \item {
% {\it
% \textbf{
% It would also have been nice to understand better whether the use of the
% epoch model (versus not) has an essential impact on tree reconstruction.
% }% END: textbf
% }% END: it
% }% END: item
% \\
% We are performing a set of simulations to asses this, and the obrained results will be presented during the public defence.
% % Running (epoch\_tree\_simulation1 epoch\_tree\_simulation2)
% 
% 
% \item {
% {\it
% \textbf{
% How mislead one could be if the data were generated under an epoch model, but analyzed under a simpler model.
% }% END: textbf
% }% END: it
% }% END: item
% \\
% We are performing a set of simulations to asses this, and the obrained results will be presented during the public defence.
% 
% 
% \end{itemize}

@@

\end{itemize}

% 2
\item {
{\it
\textbf{
It is sometimes unclear in Chapter 2 whether the analysis is performed under a fixed evolutionary tree or not. 
This is especially so in the evaluation of Bayes Factors in Table 2.3. 
It would also have been nice to understand better whether the use of the epoch model (versus not) has an essential impact on tree reconstruction. 
Somewhat related, it would have been nice to learn from the simulation study how mislead one could be if the data were generated under an epoch model, but analyzed under a simpler model. 
I am not asking for additional analyses here, but merely for additional clarification or insight.
}% END: textbf
}% END: it
}% END: item
\\
% FB: I don't get the first part of this comment. In the Intro we clearly write:
% "Our Bayesian approach also does not condition on a fixed tree topology but averages over all plausible evolutionary histories." (p 46) \\
% PL: OK, but to make it even more explicit I have added this to the results section (in simulation results): 'In all our analyses of the replicate data, as well as in the analyses of the real-world data sets in the next sections, we consider the tree as random and estimate it from the nucleotide data.' (this still needs to be tagged).
To re-iterate the fact that our model does not condition on a known tree topology we now write:

\begin{quote}
\myeditsvtwo
\end{quote}

We now also entertain a number of simulations which asses the impact of correcting for time-heterogeneity on tree reconstruction. 
The simulations which address the second part of this comment are now being performed and the obtained results can serve as a basis for discussion following the public defence.

@@

%%%%%%%%%%%%%%%%%%%%
%--- COMMENTS---%
%%%%%%%%%%%%%%%%%%%%
\section*{Comments by David Posada}

\begin{itemize}

% 1
\item {
{\it
\textbf{
I missed a brief primer of Bayesian model selection in the introductory chapter.
Techniques such Bayes factors, path sampling or stepping-stone sampling are used throughout but never explained. 
This should be very easy to fix.
}% END: textbf
}% END: it
}% END: item
\\
There is now a subsection called \emph{Bayesian model selection} at the end of Chapter 1, which gives a primer on Bayesian model selection that should be sufficient to follow the rest of the manuscript.

% 2
\item {
{\it
\textbf{
In the simulations in Chapter 2 you assume you know the true epoch times, but this information is rarely available in real life. 
Simulation 1 addresses the impact of model misspecification partially, as the true process does not have epochs. 
It could be a good idea to further check the impact of model misspecification, in particular simulating a number of epochs that do not correspond with the assumed ones (you later mention this idea in the final chapter but do not end up testing its potential biasing effect). 
Also, simulating under a non-epoch model - for example where different process can apply to contemporaneous branches, would be fundamental to understand the robustness of the model. 
This does not need to be done before the defense, but it could be a nice extension for the future.
}% END: textbf
}% END: it
}% END: item
\\
We are performing a set of simulations to asses this, and the obtained results can serve as a basis for discussion following the public defence.

@@

% 3
\item {
{\it
\textbf{
In general, there is some repetitive text, sometimes with identical phrases/paragraphs, between the introduction and some of the other chapters. 
On the other hand, some relevant aspects of the research are discussed in the general discussion in Chapter 5, instead of in the corresponding `research' chapter. 
This is true in particular for Chapter 3, were a very concise description of SPREAD is provided. 
I understand these redundancies and fragmentations are a byproduct of reproducing papers as chapters, but however these issues have been corrected in some occasions. 
For example, chapter 2 is not identical to the SysBio paper. 
I suggest you try a bit to further consolidate the different chapters.
}% END: textbf
}% END: it
}% END: item
\\
We carefully screened for repeated paragraphs between the introduction and the other chapters, and could only identify one paragraph that is repeated at th ebeginning of Subsection~\ref{sub:subst_models}.
This paragraph is now re-sentenced to remove the repetition.
We acknowledge that isolated sentence might still carry some degree of resemblance, yet we preffer not to remove this in order to also allow the subsequent chapter to function as self-contained work, for readers possibly interested only in a part of the presented work.

% 4
\item {
{\it
\textbf{
The simulation software piBUSS described in Chapter 4 can be very useful, in particular in regards to its connectivity with BEAST. 
The possibility of simulating discrete phylogeographic traits is very convenient. 
Here it would have been interesting to see, for the sake of comparison and interpretation, the correspondence between simulated and estimated tMRCAs when the true model has one epoch with omega=0.5 (or even better, omega equal to the weighted mean of the n-epoch model). 
This does not need to be done before the defense, but it could be a nice extension for the future.
}% END: textbf
}% END: it
}% END: item
\\
We are performing a set of simulations to asses this, and the obtained results can serve as a basis for discussion following the public defence.

@@

\begin{quote}
\myeditDPfive
\end{quote}

% 6
\item {
{\it
\textbf{
Page 25, 31: likelihood should be $P(X| \theta)$, not $P(\theta | X)$.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 7
\item {
{\it
\textbf{
Page 34: on the left-term of equation 1.37, the denominator should be
$P(\theta[t]|X) q(\theta^{*} | \theta[t])$
not 
$P(\theta[t]|X) q(\theta[t] | \theta^{*})$
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 8
\item {
{\it
\textbf{
Page 52: according to the legend in Figure 2.3, a two-epoch plot is missed.
}% END: textbf
}% END: it
}% END: item
\\
Entire set up for the employed simulations is presented in one plot by the use of alternating shading. 
This was perhaps not clear enough from the Figures caption, I have thus rephrased it to state explicitely:

### 2015-02-11 Response.tex

\begin{itemize}

% 1
\item {
{\it
\textbf{
The thesis starts with a very nice introduction, which I think is very insightful for the reader who is not familiar with phylogenetic and/or Bayesian inference. 
In spite of this, I believe that the readability of the later chapters would be further improved if the introduction:
}% END: textbf
}% END: it
}% END: item

\begin{itemize}

% 1a
\item {
{\it
\textbf{
Briefly described the evolutionary models, such as GY94, that are later used.
}% END: textbf
}% END: it
}% END: item
\\
I have now added a \emph{Markov models of codon evolution} subsection in the introductory chapter, which discusses the most popular codon model implementations (see Subsection ~\ref{sub:codon}).

% 1b
\item {
{\it
\textbf{
Briefly described how to accommodate time-stamped sequences, which is now considered in the analyses in Chapter 2, but not really discussed.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, see Subsection~\ref{sub:dated_trees}, which describes the methods used in BEAST to accommodate time-stamped sequences as a source of calibration for trees.
%
%PL: TODO: rewrite the final paragraph to reflect the dated-tip likelihood model: Essential, we now consider heights (instead of branch lengths) and introduce a new parameter, the rate of evolution r, which allows rescaling genetic distance into time. The task is now to estimate the heights and r from the sequence data, keeping the ti[ heights proportional to their sampling time, with the constant of proportionality being our estimable r.
%

### 2015-02-11 chapters/chapter01.tex
revision edits

Time-stamped sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur, can be used to calibrate the molecular clock and convert branch lengths into actual time units.

Fo simplicity let us assume a single site, where character sequence data was sampled on dates ranging from $1983$ to $2000$, as seen in Figure~\ref{fig:timing_sequences}.
We also assume that the tree topology is known and we want to estimate the unknown times $T_{0},T_{6},T_{7},T_{8}$.

\begin{figure}[h!]
\centering
\includegraphics[scale=0.4]{timing_sequences} 
\caption{
{ \footnotesize 
{\bf Dated phylogenetic tree.} 
Discrete characters $CCACT$ observed at the external nodes (the tips of the tree) were sampled at different points in time.
The tree topology is known and assumed to be true.
The unknown dates of the internal nodes are labelled $T_{0},T_{6},T_{7},T_{8}$.
}% END: footnotesize
}
\label{fig:timing_sequences}
\end{figure}

>>>>

Time-stamped or `heterchronous' sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur between the sampling times, can be used to calibrate the molecular clock and convert branch lengths into actual time units.

Fo simplicity let us assume a single site, where character sequence data was sampled on dates ranging from $1983$ to $2000$, as seen in Figure~\ref{fig:timing_sequences}.
We also assume that the rooted tree topology is known and we want to estimate the unknown times $T_{0},T_{6},T_{7},T_{8}$.

\begin{figure}[h!]
\centering
\includegraphics[scale=0.4]{timing_sequences} 
\caption{
{ \footnotesize 
{\bf Dated phylogenetic tree.} 
Discrete characters $CCACT$ observed at the external nodes (the tips of the tree) were sampled at different points in time.
The tree topology is assumed to be known without error, and also the root is known.
The unknown dates of the internal nodes are labelled $T_{0},T_{6},T_{7},T_{8}$.
}% END: footnotesize
}
\label{fig:timing_sequences}
\end{figure}
